---
title: "Unit 1 - DATA 607"
author: "Marc Campo"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Set Options

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=FALSE, message=FALSE, warning=FALSE)
```

# APPROACH

## *Introduction*

The data source is a simulated data set that I based on research we
conducted examining the relationship between cognition and mobility
gains in patient rehabilitation for stroke. For IRB and HIPPA reasons, I
could not post the actual data set but the results and structure are
similar though our original data includes over 1,000 variables.

The approach was to post the data on GitHub and then clean it in R
Markdown as well as polish the names and value labels.

# CODE

## *Import Data*

Data were imported from GitHub as a CSV file.

```{r}
url <- "https://raw.githubusercontent.com/cam5671/DATA-607/main/unit1/Mobility%20Gains%20in%20Stroke%20Rehabilitation.csv"
data <- read.csv(url)
head(data)
```

## *Drop Unnecessary Columns*

The purpose is to examine the effect of admissions factors (age, sex,
and clinical variables) on the Mobility Score at discharge. The mobility
score is part of section GG data for inpatient rehabilitation and
collected on every patient across the US. The predictors include
Cognitive Score as assessed by the Montreal Cognitive Assessment, Lower
Extremity Strength as measured by the Motricity Index, Age, Sex, and
Length of Stay. Mobility score at discharge is the outcome. The most
important predictor is likely the mobility score at admission which is
also included. Discharge values for cognition and strength as well as
hospital dates and clinic notes are unnecessary and are omitted.

```{r}
data <- data[,-c(2:4, 9, 11)]
names(data)
```

## Change Variable Names

Variable names can be improved to be more presentable and legible.
Changing names can be perilous in R if you use a vector of names and
assignment them in order. I will change the names one at a time.

The final data set will include:

-   ID: Participant ID
-   Sex: Sex
-   LOS: Length of Stay in Inpatient Rehabilitation
-   Cognition: Cognitive Score on Montreal Cognitive Assessment
-   MobilityADM: Mobility Score on Admission (Section GG of Inpatient Rehabilitation Assessment)
-   MobilityDC: Mobility Score at Discharge

```{r}
colnames(data)[1] <- 'ID'
colnames(data)[2] <- 'Age'
colnames(data)[3] <- 'Sex'
colnames(data)[4] <- 'LOS'
colnames(data)[5] <- 'LEStrength'
colnames(data)[6] <- 'Cognition'
colnames(data)[7] <- 'MobilityADM'
colnames(data)[8] <- 'MobilityDC'
head(data)
```

## *Value Labels*

Sex is coded as 0 for Male, 1 for Female. The CSV file, however, did not
include 'value labels' that we would typically see in STATA or SPSS or
SAS. We can convert it in R to a factor variable.

```{r}
data$Sex <- factor(data$Sex, levels = 0:1, labels=c('Male', 'Female'))
table(data$Sex)
```

## Quick Screen of the Data

The data set is now established and clean. There are no missing values.
The following is a quick screen of the data prior to what the main
analysis was (linear regression though many other approaches are also
possible). We already have a table for our categorical variable so plots
for continuous variable will work here. Plots that show the data points,
the medians and quartiles, as well as means provide the reader with a
full picture of central tendency, skewness and outliers. The red dot is
the mean superimposed on the boxplot. The variables are relatively
normally distributed and you can see an increase in mobility from
admission to discharge. When we simulate data using rnorm, the data end
up artificially close normality. In our real data we see more skew with
variables like Length of Stay (LOS).

```{r}
library(ggplot2)
library(gridExtra)
library(ggplot2)
library(gridExtra)

plotf <- function(x) {

  ggplot(data, aes(x = "", y = .data[[x]])) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(
      width = .05,
      color = "steelblue",
      pch = 1
    ) +
    stat_summary(
      fun = mean,
      geom = "point",
      pch = 16,
      color = "red",
      size = 5
    ) +
    labs(
      title = x,
      x = NULL,
      y = NULL
    )
}

vars <- c(
  "Age",
  "LOS",
  "LEStrength",
  "Cognition",
  "MobilityADM",
  "MobilityDC"
)

plots <- lapply(vars, plotf)

grid.arrange(
  grobs = plots,
  ncol = 3
)
```

## *Descriptive Statistics for Variables*

The plots provide a full picture but descriptives can help as well.

```{r}
library(psych)
library(gt)
library(dplyr)


vars <- c("Age", "LOS", "LEStrength", "Cognition",
          "MobilityADM", "MobilityDC")


desc <- psych::describe(data[vars])

desc_table <- desc %>%
  mutate(
    Variable = rownames(.),
    IQR = sapply(data[vars], IQR, na.rm = TRUE),
    Range = paste0(round(min, 1), "–", round(max, 1))
  ) %>%
  select(
    Variable,
    N = n,
    Mean = mean,
    SD = sd,
    Median = median,
    IQR,
    Range
  )

desc_table %>%
  gt() %>%
  fmt_number(
    columns = c(Mean, SD, Median, IQR),
    decimals = 1
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  cols_align(
    align = "left",
    columns = Variable
  ) %>%
  tab_header(
    title = "Descriptive Statistics"
  )
```

## Conclusion

The data are ready for analysis and are appropriate for multiple linear
regression though these data (and the real data) violate the assumption
of heteroscedasticity and dominance analysis was used to determine the
relative importance of the predictors. Other techniques like machine
learning might offer better predictive accuracy and would be suitable
for use on this dataset.

## Acknowledgments

Generative AI (ChatGPT, OpenAI) was used for guidance on working with
GitHub. It was also used to refine the loop used for the plots and it
was used to format the table of descriptive statistics to improve the
appearance from the standard output in the Psych package.
